Thermal reactionomes reveal adaptive responses to thermal extremes in warm

### 2014-10-24 doc/PNAS_cover_letter.tex
final version PNAS cover letter

\fancypagestyle{first}{%
  \fancyhf{}
  \setlength{\headheight}{40pt}
  \fancyhead[L]{\includegraphics[scale=0.1]{UVM_Official_Logo.jpg}} 
  \fancyhead[R]{\small Department of Biology \\102  Marsh Life Sciences \\ Burlington, VT 05405 }
  \renewcommand{\headrulewidth}{0.4pt}
  \renewcommand{\footrulewidth}{0pt}}
\pagestyle{first}

@@

We are excited to submit our manuscript \emph{Investigation of thermal
reactionomes reveals differing responses to thermal extremes in warm and
cold-climate ant species} for consideration as a Research Report in
\emph{The Proceedings of the National Academy of Sciences}. There are
several aspects of the work that we think make it worthy of publication
in \emph{PNAS}.

In face of contemporary climate change, many species are predicted to be at 
risk of extinction. However, there is also evidence that species can physiologically
acclimate and rapidly adapt to changing environmental conditions, potentially
rescuing them from extinction. This paradox poses a dilemma for evolutionary
ecologists. In this study, we evaluate plasticity and adaptive differentiation
in gene expression of closely related cool and warm-climate ant species.

We present a novel analysis by extending the reaction
norm approach to the study of gene expression. We define the
\emph{reactionome} as a characterization of the reaction norm for all
genes in an organism's transcriptome across an environmental gradient. 
We believe that this approach is an important contribution as it allows us
to make quantitative comparisons of plasticity in gene expression between
species. Importantly, this analysis does not depend on the functional 
annotation of the genes, and thus is robust to criticisms of "storytelling" 
in RNAseq studies based on annotation of differentially expressed 
genes (Pavlidis et al. 2012 MBE).

By applying the reactionome approach to two related ant species,
\emph{Aphaenogaster picea} and \emph{A. carolinensis}, from colder and
warmer climates, respectively, we gain insight on the relationship
between short-term physiological acclimation and longer-term adaptive
change in response to thermal extremes. In particular, we find that
adaptation to chronic high temperatures involves different mechanisms
than those used to cope with rarely encountered stressors, which may
limit adaptive potential of northern species in the face of rapidly
shifting climatices.

@@

\noindent{Thank you for consideration of our work,} \\
\includegraphics[scale=1]{JSG_signature.png}  \\

Dr. John Stanton-Geddes

>>>>

\fancypagestyle{first}{%
  \fancyhf{}
  \setlength{\headheight}{30pt}
  \fancyhead[L]{\includegraphics[scale=0.1]{UVM_Official_Logo.jpg}} 
  \fancyhead[R]{\small Department of Biology \\102  Marsh Life Sciences \\ Burlington, VT 05405 }
  \renewcommand{\headrulewidth}{0.3pt}
  \renewcommand{\footrulewidth}{0pt}}
\pagestyle{first}

@@

We are excited to submit our manuscript \emph{Thermal
reactionomes reveal adaptive responses to thermal extremes in warm and
cold-climate ant species} for consideration as a Research Report in
\emph{The Proceedings of the National Academy of Sciences}. There are
several aspects of the work that we think make it worthy of publication
in \emph{PNAS}.

In face of contemporary climate change, many species are predicted to be at 
risk of extinction. However, there is also evidence that species can physiologically
acclimate and rapidly adapt to changing environmental conditions, potentially
rescuing them from extinction. How these two processes are related at the molecular
level is poorly understood, hampered by the limitations inherent in extrapolating 
complex responses and functional interactions from the relatively simple experimental
design of most gene expression experiments to date.

In this study, we resolve this problem by extending the reaction
norm approach to the study of gene expression. We define the
\emph{reactionome} as a characterization of the reaction norm for all
genes in an organism's transcriptome across an environmental gradient. 
We believe that this approach is an important contribution as it allows us
to make quantitative comparisons of plasticity in gene expression between
species. Importantly, this analysis does not depend on the functional 
annotation of the genes, and thus is robust to criticisms of "storytelling" 
in RNAseq studies based on annotation of differentially expressed 
genes (Pavlidis et al. 2012 MBE).

By applying the reactionome approach to two related ant species,
\emph{Aphaenogaster picea} and \emph{A. carolinensis}, from colder and
warmer climates, respectively, we gain insight on the relationship
between short-term physiological acclimation and longer-term adaptive
change in response to thermal extremes. In particular, we find that while
extensions of lower critical thermal limits involve enhancing existing plasticity,
adaptation to chronic high temperatures involves a shift from induced resistance
toward novel tolerance mechanisms, which may limit adaptive potential 
in the face of rapidly rising temperatures associated with global climate change.

@@

\noindent{Thank you for consideration of our work,} \\
\includegraphics[scale=0.8]{JSG_signature.png}  \\
Dr. John Stanton-Geddes on behalf of all the authors
